

## DEGs Identification: Pseudobulk
### Summarize to Pseudobulk
```{r ST-pseudobulk-DEGs, eval=FALSE, include=FALSE}
source("script/load_checkpoint.r")
#---------------------------------------------------------------
library(SingleCellExperiment)
library(Seurat)
# Split by brain region
meta <- data.frame(testcombo@cell_metadata$cell$rna@metaDT)
row.names(meta) <- meta$cell_ID
gobj.count <- CreateSeuratObject(testcombo@expression$cell$rna$raw@exprMat,
                              meta.data=meta)
sce <- as.SingleCellExperiment(gobj.count)
summed <- scuttle::aggregateAcrossCells(sce,
             id=colData(sce)[,c("cell_types", "list_ID")])
saveRDS(summed, "001Giotto_out/003DEGs/DESeq/gobj.pseudocount_by_sample_br.rds")
summed.whole <- scuttle::aggregateAcrossCells(sce,
             id=colData(sce)[,c("list_ID")])
saveRDS(summed.whole, "001Giotto_out/003DEGs/DESeq/gobj.pseudocount_by_sample.rds")
```

### DESeq2 (old)
```{r}
# -----------------------------------
# PCA for each sample + brain region
mtx <- assays(summed)$counts
meta <- colData(summed)
row.names(meta) <- paste0(meta$list_ID, "_", meta$cell_types)
colnames(mtx) <- paste0(meta$list_ID, "_", meta$cell_types)
sobj.agg.L1 <- CreateSeuratObject(mtx, meta.data=as.data.frame(meta))
sobj.agg.L1 <- NormalizeData(sobj.agg.L1)
sobj.agg.L1 <- FindVariableFeatures(sobj.agg.L1)
sobj.agg.L1 <- ScaleData(sobj.agg.L1)
sobj.agg.L1 <- RunPCA(sobj.agg.L1)
p1 <- DimPlot(sobj.agg.L1, reduction="pca", shape.by="batch", group.by="cell_types")
p2 <- DimPlot(sobj.agg.L1, reduction="pca", shape.by="batch", group.by="treatment")
dir.create("001Giotto_out/003DEGs/DESeq")
ggplot2::ggsave("001Giotto_out/003DEGs/DESeq/test.png", p1)
ggplot2::ggsave("001Giotto_out/003DEGs/DESeq/test2.png", p2)
#----------------------------------------
# PCA for each sample
sce.agg_by_sample <- scuttle::aggregateAcrossCells(sce,
             id=colData(sce)[,"list_ID"])
mtx <- assays(sce.agg_by_sample)$counts
meta <- colData(sce.agg_by_sample)
row.names(meta) <- meta$list_ID
colnames(mtx) <- meta$list_ID
sobj.agg.L1 <- CreateSeuratObject(mtx, meta.data=as.data.frame(meta))
sobj.agg.L1 <- NormalizeData(sobj.agg.L1)
sobj.agg.L1 <- FindVariableFeatures(sobj.agg.L1)
sobj.agg.L1 <- ScaleData(sobj.agg.L1)
sobj.agg.L1 <- RunPCA(sobj.agg.L1, npcs=2)
p1 <- DimPlot(sobj.agg.L1, reduction="pca", shape.by="batch", group.by="treatment")
ggplot2::ggsave("001Giotto_out/003DEGs/DESeq/PCA.each_sample.png", p1)
# -----------------------------------
# Differential Expression Analysis
# dhlab
if (F) {
  # Old code.
  library(DESeq2)
 
  pair_wise_result <- list()
 
  # If only treatment is used, there are no results. If treatment + batch is used, there are more results.
  for (ct in unique(summed$cell_types)){
    if (ct=="C12") next
    # ct <- "Hippocampus"
    summed.sub <- subset(summed, cell_types==ct)
    dds <- DESeqDataSet(summed.sub, ~ treatment + batch)
    dds <- DESeq(dds)
    # resultsNames(dds)
    res <- results(dds, name = "treatment_PLX5622_vs_FADCtrl")
    res.df <- data.frame(res)
    res.df <- res.df[!is.na(res.df$padj), ]
    res.df <- subset(res.df, padj < 0.05)
    pair_wise_result[[ct]] <- res.df
  }
 
  saveRDS(pair_wise_result, "001Giotto_out/003DEGs/DESeq/pair_wise_result.rds")
 
  for (x in names(pair_wise_result)){
    df <- pair_wise_result[[x]]
    if (nrow(df) == 0) {
      print(x)
      pair_wise_result$x <- NULL
      next
    }
    df$gene <- row.names(df)
    df$brain_region <- x
    pair_wise_result[[x]] <- df
  }
 
  st.DEGs.deseq <- Reduce(rbind, pair_wise_result)
  row.names(st.DEGs.deseq) <- 1:nrow(st.DEGs.deseq)
  write.csv(st.DEGs.deseq, "001Giotto_out/003DEGs/DESeq/st.DEGs.deseq.csv")
}
# Try using only the first two batches of single cells for differential expression analysis.
# Investigate genes that are significant in both methods one by one.
# DEGs in cell types highly expressed in 5xFAD?
# Without separating brain regions
meta <- data.frame(testcombo@cell_metadata$cell$rna@metaDT)
row.names(meta) <- meta$cell_ID
gobj.count <- CreateSeuratObject(testcombo@expression$cell$rna$raw@exprMat,
                              meta.data=meta)
sce <- as.SingleCellExperiment(gobj.count)
summed <- scuttle::aggregateAcrossCells(sce,
             id=colData(sce)[,"list_ID"])
dds <- DESeqDataSet(summed, ~ treatment + batch)
dds <- DESeq(dds)
# resultsNames(dds)
res <- results(dds, name = "treatment_PLX5622_vs_FADCtrl")
res <- data.frame(filt.res(res))
write.csv(res, "001Giotto_out/003DEGs/DESeq/whole_brain.csv")
counts.normalized <- counts(dds, normalized=T)
up.genes <- res[res$log2FoldChange > 0, ]
 # [1] "mt-Atp8" "mt-Nd4l" "mt-Nd3" "mt-Nd2" "Nr1d1": oxidative phosphorylation
# "Dbp" "Slc13a3" "Rps29"
 # [8] "Polr2l" "Slc13a4" "Arc" "Ciart"
# Dbp: transcription factor family. Binds to genes such as albumin, CYP2A4, and CYP2A5. Involved in the regulation of some circadian rhythm genes.
# Nr1d1: This gene encodes a transcription factor that is a member of the nuclear receptor subfamily 1. In particular, this protein represses the circadian clock transcription factor aryl hydrocarbon receptor nuclear translocator-like protein 1 (ARNTL).
# Ciart: behavior. Circadian regulation of gene expression. Transcription factor, negative regulation.
# Arc: regulation of dendritic spine morphogenesis. Regulation of cell morphogenesis.
# Polr2l: Signaling by Receptor Tyrosine Kinases
# Slc11a1|Slc13a3|Slc13a4: Transport of bile salts and organic acids, metal ions and amine compounds
# Metascape enrichment analysis.
# After enrichment analysis, many similar pathways appear. See if these pathways can be divided into relatively independent modules.
# Take genes from the top 10 pathways and calculate intersection indices.
# Remove genes from the top 10 pathways from all DEGs and see what remains.
all.list <- read.csv("001Giotto_out/003DEGs/DESeq/whole_brain.")
top10.genes <- unique(unlist(strsplit(all.list$Hits[1:10], "|", fixed=T)))
other.genes <- setdiff(row.names(res), c(top10.genes, row.names(up.genes)))
# [1] "Fcrls" "Tmem119" "P2ry12" "Olfml3" "Crybb1" "Siglech" "Cst7"
# [8] "Gngt2" "Rnase4" "Gpr34" "Ltc4s" "Itgb5" "Ptpn18" "P2ry13"
# [15] "Abi3" "Slco2b1" "Hexa" "Plxdc2" "Lpcat2" "Eif2s3y" "Fermt3"
# [22] "Ptgs1" "Lgmn" "Tbxas1" "Cyth4" "Shank3" "F11r" "Phyhd1"
# [29] "Lzts3"
for (i in other.genes) {
  print(i)
  print(head(all.list[grepl(i, all.list$Hits), "Description"]))
  cat("\n")
}
# No pathways were enriched.
# Immune-related:
# Fcrls, Olfml3, Gpr34, Slco2b1, Plxdc2, Eif2s3y, Cyth4, Phyhd1:
#
# Tmem119: Function unknown. Positive regulation of phosphorylation?
# P2ry12: Related to cell migration
# Immune-related: signal transduction, transcriptional activation, chemotaxis, motility, phagocytosis, degradation, antigen presentation.
# Cst7: regulation of inflammatory response
# Rnase4: defense response to bacterium
# Itgb5: Adaptive Immune System
# Siglech: endocytosis
# Hexa: Lysosome
# Lpcat2: Metabolism of lipids
# Gngt2: Chemokine signaling pathway
# Fermt3, F11r: leukocyte cell-cell adhesion
# Lgmn: antigen processing and presentation
# Signal regulation:
# Ptpn18: Signaling by Receptor Tyrosine Kinases
# P2ry13: G alpha (i) signalling events
# Development-related:
# Crybb1: visual system development
# Tbxas1, Ptgs1, Ltc4s: Eicosanoid synthesis, Platelet activation
# Neural-related:
# Shank3: locomotion, regulation of long-term synaptic potentiation
# Lzts3: regulation of postsynapse organization
```

### DESeq2
```{r}
library(DESeq2)
summed <- readRDS("../001ST/001Giotto_out/003DEGs/DESeq/gobj.pseudocount_by_sample_br.rds")
for (ct in unique(summed$cell_types)){
  if (ct=="C12") next
  # ct <- "Hippocampus"
  summed.sub <- subset(summed, ,cell_types==ct)
  dds <- DESeqDataSet(summed.sub, ~ treatment + batch)
 
  levels(dds$treatment)
  # [1] "FADCtrl" "PLX5622" "WTCtrl"
 
  dds <- DESeq(dds)
  saveRDS(dds, paste0("../001ST/001Giotto_out/003DEGs/DESeq/", make.names(ct), ".rds"))
}
result.list <- list()
for (ct in unique(summed$cell_types)){
  if (ct=="C12") next
  # resultsNames(dds)
  in.path <- paste0("../001ST/001Giotto_out/003DEGs/DESeq/", make.names(ct), ".rds")
  dds <- readRDS(in.path)
  res <- results(dds, name = "treatment_PLX5622_vs_FADCtrl")
  res.df1 <- data.frame(res)
  res.df1$gene <- row.names(res.df1) # If the gene is used as row name, gene names may cause errors when merging tables with duplicate genes.
  res.df1$comparison <- "PLX5622_vs_FADCtrl"
 
  res <- results(dds, contrast = c("treatment", "FADCtrl", "WTCtrl"))
  res.df2 <- data.frame(res)
  res.df2$comparison <- "FADCtrl_vs_WTCtrl"
  res.df2$gene <- row.names(res.df2)
  result.list[[ct]] <- rbind(res.df1, res.df2)
}
saveRDS(result.list, "../001ST/001Giotto_out/003DEGs/DESeq/DESeq_result.sample_br.list.rds")
if (F) {
  # Unfiltered tables take up space and are rarely viewed, so it is better to save as rds.
  result.df <- Reduce(rbind, result.list)
  write.csv(result.df, "../001ST/001Giotto_out/003DEGs/DESeq/DESeq.sample_br.csv")
}
```

### Pseudobulk DEGs Heatmap
```{r}
result.list <- readRDS("../001ST/001Giotto_out/003DEGs/DESeq/DESeq_result.sample_br.list.rds")
df.merged <- data.frame()
for (BR in names(result.list)) {
  # BR <- names(result.list)[1]
  if (BR %in% c("Third Ventricle", "Lateral Ventricle")) next
  df <- result.list[[BR]]
  df.sub <- subset(df, padj < 0.05 & baseMean > 10 & abs(log2FoldChange) > 0.4)
  df.sub$Brain_region <- BR
  df.merged <- rbind(df.merged, df.sub)
}
write.csv(df.merged, "../001ST/001Giotto_out/003DEGs/DESeq/DESeq_result.sample_br.DEGs.csv")
# df.merged <- read.csv("../001ST/001Giotto_out/003DEGs/DESeq/DESeq_result.sample_br.DEGs.csv", row.names=1)
df.merged$Brain_region.uniq <- paste0(df.merged$Brain_region,
                                      ifelse(df.merged$comparison == "FADCtrl_vs_WTCtrl", "1", "2"))
library(reshape2)
df.merged.sub <- subset(df.merged, comparison == "FADCtrl_vs_WTCtrl")
df.dcast <- dcast(gene~Brain_region, value.var = "log2FoldChange", data=df.merged)
df.dcast[is.na(df.dcast)] <- 0
df.dcast <- data.frame(df.dcast[, -1], row.names=df.dcast$gene)
p <- pheatmap(df.dcast, cluster_cols = T, cutree_rows = 5, scale="row")
```

#### AD vs WT
In transcriptome analysis, I want to use heatmaps to cluster genes based on their expression patterns across different treatments and display the genes contained in each cluster. Using `cutree_rows` in pheatmap can accomplish this task, but now I encounter a problem: there are too many genes, and the heatmap cannot display them completely. What are the solutions?

I want to perform clustering with pheatmap and display all genes, but without showing row names. Then put the genes of each cluster after clustering into a table. How can this be done in R?
```{r}
df.merged.sub <- subset(df.merged, comparison == "FADCtrl_vs_WTCtrl")
df.dcast <- dcast(gene~Brain_region, value.var = "log2FoldChange", data=df.merged.sub)
df.dcast[is.na(df.dcast)] <- 0
df.dcast <- data.frame(df.dcast[, -1], row.names=df.dcast$gene)
mat_scaled <- df.dcast
for (i in 1:nrow(mat_scaled)) {
  row_abs_max <- max(abs(mat_scaled[i, ]), na.rm = TRUE)
  if (row_abs_max > 0) { # Avoid division by zero
    mat_scaled[i, ] <- mat_scaled[i, ] / row_abs_max
  }
  # If the entire row is 0, it remains 0
}
# Step 2: Define colors and breaks to ensure 0 is white, >0 is red, <0 is blue
my_colors <- colorRampPalette(c("navy", "white", "firebrick3"))(100)
# breaks: evenly distributed from -1 to 1, ensuring 0 is in the middle (white)
my_breaks <- seq(-1, 1, length.out = 101)
k <- 6
p <- pheatmap(mat_scaled, scale="none", show_rownames = FALSE)
row_clusters <- cutree(p$tree_row, k = k)
annotation_row <- data.frame(
  Cluster = factor(row_clusters, levels = 1:k, labels = paste0("C", 1:k))
)
rownames(annotation_row) <- rownames(mat)
p <- pheatmap(mat_scaled,
         cluster_rows = p$tree_row,
         scale = "none",
         show_rownames = FALSE,
         color = my_colors,
         breaks = my_breaks,
         annotation_row = annotation_row,
         main = "pseudobulk ST DEGs (AD vs WT)")
library(grid)
pdf("../001ST/001Giotto_out/003DEGs/DESeq/pseudobulk_ST_DEGs_AD_vs_WT.pdf", width=4.97, height=9.08)
grid.newpage()
grid.draw(p$gtable)
dev.off()
png("../001ST/001Giotto_out/003DEGs/DESeq/pseudobulk_ST_DEGs_AD_vs_WT.png", width=4.97, height=9.08, units="in", res=300)
grid.newpage()
grid.draw(p$gtable)
dev.off()
library(dplyr)
# Export table
df <- data.frame(Gene = rownames(df.dcast), Cluster = paste0("C", row_clusters)) %>%
  arrange(Cluster, Gene)
df <- aggregate(Gene~Cluster, FUN=function(x){paste(x, collapse=", ")}, data=df)
writeLines(paste0(df$Cluster, ": ", df$Gene))
# output: "C:\Users\a8799\Desktop\project\2022-10-14_mouse_spatial\doc\2026-01-13_oxidative_phosphorylation.pptx"
```

#### PL vs AD
You can refer to the heatmap on the shinyapp.

I now have a matrix: each row is a gene, each column is a brain region, and each value represents the log2(fold change) of that gene in the specific brain region.

Now I need to draw a heatmap that meets the following requirements:
1. Scale by row, while ensuring that values originally >0 remain >0 and values originally <0 remain <0 after scaling.
2. Red represents >0 (upregulated), white represents =0, blue represents <0 (downregulated).
3. Cluster by row.

Please help me write the R code.
```{r}
df.merged.sub <- subset(df.merged, comparison == "PLX5622_vs_FADCtrl")
df.dcast <- dcast(gene~Brain_region, value.var = "log2FoldChange", data=df.merged.sub)
df.dcast[is.na(df.dcast)] <- 0
df.dcast <- data.frame(df.dcast[, -1], row.names=df.dcast$gene)
# Load package
library(pheatmap)
# Assume your matrix is called mat (rows = genes, columns = brain regions, values = log2(FC))
# Step 1: Scale by row but keep the sign unchanged (by dividing by the row's maximum absolute value)
# This maps each row's values to the [-1,1] range while preserving the sign
mat_scaled <- df.dcast # Copy the original matrix
for (i in 1:nrow(mat_scaled)) {
  row_abs_max <- max(abs(mat_scaled[i, ]), na.rm = TRUE)
  if (row_abs_max > 0) { # Avoid division by zero
    mat_scaled[i, ] <- mat_scaled[i, ] / row_abs_max
  }
  # If the entire row is 0, it remains 0
}
# Step 2: Define colors and breaks to ensure 0 is white, >0 is red, <0 is blue
my_colors <- colorRampPalette(c("navy", "white", "firebrick3"))(100)
# breaks: evenly distributed from -1 to 1, ensuring 0 is in the middle (white)
my_breaks <- seq(-1, 1, length.out = 101)
# Step 3: Draw the heatmap and cluster by row
pheatmap(mat_scaled,
         scale = "none", # Already manually scaled, do not use built-in scale
         cluster_rows = TRUE, # Cluster rows
         cluster_cols = TRUE, # Can be changed to FALSE according to needs
         color = my_colors,
         breaks = my_breaks,
         main = "Heatmap of log2(FC) by Brain Regions",
         show_rownames = TRUE, # Can be changed to FALSE if there are too many genes
         show_colnames = TRUE,
         treeheight_row = 50)
```

#### Mitochondrial Gene Investigation: DEGs
```{r}
library(readxl)
df.mt <- read_excel("../001ST/000data/Mouse.MitoCarta3.0.xls", sheet = 2)
df.merged <- read.csv("../001ST/001Giotto_out/003DEGs/DESeq/DESeq_result.sample_br.DEGs.csv", row.names=1)
df.merged[df.merged$gene %in% df.mt$Symbol, c("baseMean", "log2FoldChange", "padj", "gene", "comparison", "Brain_region")]
```

### GSEA Analysis
```{r}
library(clusterProfiler)
library(fgsea)
library(msigdbr) # Used to obtain MSigDB gene sets
# My experiment has three different treatments (assumed to be a, b, c), and DESeq was used for differential expression analysis with the design parameter "~ treatment + batch". After running the DESeq function, I successfully extracted the differential expression results for c vs b using the results function. However, when I tried to extract the results for b vs a using the same method, I got the error: "Error: subscript contains invalid names". How to solve this?
dir.create("../001ST/001Giotto_out/003DEGs/GSEA")
for (BR in setdiff(unique(summed$cell_types), "C12")) {
  for (CP in c("PLX5622_vs_FADCtrl", "FADCtrl_vs_WTCtrl")) {
    res_df <- subset(result.list[[BR]], comparison == CP)
   
    nrow(res_df)
    # [1] 18551
    nrow(na.omit(res_df))
    # [1] 12614
   
    res_df <- na.omit(res_df)
   
    if (F) {
      # It seems there are not many false positives caused by low baseMean, so this step can be skipped.
      nrow(subset(res_df, baseMean > 10)) # 12614
      res_df$RANK <- rank(-res_df$baseMean)
      res_df <- subset(res_df, RANK <= 12000)
    }
   
    gene_list <- res_df$log2FoldChange
    names(gene_list) <- res_df$gene
    gene_list <- sort(gene_list, decreasing = TRUE) # From highest upregulated to highest downregulated
   
    COLLECT <- "C5"
   
    m_df <- msigdbr(species = "Mus musculus", collection = COLLECT)
   
    gsea_res4 <- GSEA(
      geneList = gene_list, # Sorted gene list
      exponent = 1, # Weight exponent, usually 1
      minGSSize = 5, # Minimum gene set size (filter out too small sets)
      maxGSSize = 500, # Maximum gene set size (filter out too large sets)
      pvalueCutoff = 0.1, # P-value cutoff
      pAdjustMethod = "BH", # FDR correction method
      TERM2GENE = data.frame(term = m_df$gs_name,
                             gene = m_df$gene_symbol) # Or directly use the gene set
      # Note: Adjust parameters according to the output of `msigdbr`
    )
   
    saveRDS(gsea_res4, paste0("../001ST/001Giotto_out/003DEGs/GSEA/", make.names(BR), ".", CP, ".gsea.rds"))
  }
}
```

### Organize GSEA Results
```{r}
# Organize GSEA results
df.list <- list()
n <- 1
for (CP in c("PLX5622_vs_FADCtrl", "FADCtrl_vs_WTCtrl")) {
  for (BR in setdiff(unique(summed$cell_types), "C12")) {
    out.path <- paste0("../001ST/001Giotto_out/003DEGs/GSEA/", make.names(BR), ".", CP, ".gsea.rds")
    gsea_result <- readRDS(out.path)
    df <- gsea_result@result # [ c("ID", "NES", "qvalue", "core_enrichment")]
   
    if (nrow(df) > 0) {
      df.cols <- colnames(df)
      df$brain_region <- BR
      df$comparison <- CP
      df <- df[, c("brain_region", "comparison", df.cols)]
      df.list[[n]] <- df
      n <- n + 1
    }
    # p <- gseaplot2(gsea_result, 1:5, pvalue_table = T, pvalue_table_columns = "pvalue")
    # ggsave("")
  }
}
df.merged <- Reduce(rbind, df.list)
df.merged <- df.merged[order(df.merged$brain_region,
                   df.merged$comparison,
                   df.merged$qvalue), ]
out.path <- paste0("../001ST/001Giotto_out/003DEGs/GSEA/st.gsea.csv")
write.csv(df.merged, out.path)
```

### Identify Rescued Pathways
#### Load some functions.
```{r}
add_genelist <- function(df) {
  gl.vec <- c()
  for (i in 1:nrow(df)) {
    BR1 <- df[i, "brain_region"]
    ID1 <- df[i, "ID"]
    df.sub <- subset(df.merged, ID == ID1 & brain_region == BR1)
    gl <- unique(unlist(strsplit(df.sub$core_enrichment, "/")))
    gl.vec[i] <- paste(gl, collapse = ",")
  }
  return (gl.vec)
}
keep_rows_contain_any_genes <- function(in.df, in.gl, gl.col="genelist", gl.sep=",") {
  keep.row <- sapply(in.df[[gl.col]], function(x) {
    gl <- strsplit(x, gl.sep)[[1]]
    any(gl %in% in.gl)
  })
  return(in.df[keep.row, ])
}
exclude_rows_contain_any_genes <- function(in.df, in.gl) {
  keep.row <- sapply(in.df$genelist, function(x) {
    gl <- strsplit(x, ",")[[1]]
    !any(gl %in% in.gl)
  })
  return(in.df[keep.row, ])
}
add_intersect_col <- function(in.df, in.gl, gl.col="genelist", gl.sep=",") {
  in.df$intersect.genes <- sapply(in.df[[gl.col]], function(x) {
    gl <- strsplit(x, gl.sep)[[1]]
    paste(intersect(gl, in.gl), collapse=gl.sep)
  })
  return(in.df)
}
get_rows_contain_genes_in_df.merged <- function(df.merged, gene.list) {
  df <- keep_rows_contain_any_genes(df.merged, in.gl = gene.list,
                            gl.col = "core_enrichment", gl.sep = "/")
  df <- add_intersect_col(df, gene.list, gl.col = "core_enrichment", gl.sep="/")
  return(df)
}
```

#### Identify Up-regulated Rescued Pathways
```{r}
library(reshape2)
df.merged <- read.csv("../001ST/001Giotto_out/003DEGs/GSEA/st.gsea.csv", row.names=1)
df.wide <- dcast(ID+brain_region~comparison, value.var = "NES", data=df.merged)
df.wide.sub <- na.omit(df.wide)
# Up-rescued pathways?
df.wide.sub.up_rescued <- subset(df.wide.sub, FADCtrl_vs_WTCtrl < 0 & PLX5622_vs_FADCtrl > 0)
# Add gene names
# View the genes contained in pathways of interest
df.wide.sub.up_rescued$genelist <- add_genelist(df.wide.sub.up_rescued)
# View(df.wide.sub.up_rescued[order(df.wide.sub.up_rescued$brain_region), ])
subset(df.wide.sub.up_rescued, ID == "GOCC_RESPIRATORY_CHAIN_COMPLEX")
subset(df.wide.sub.up_rescued, ID == "GOBP_PROTON_MOTIVE_FORCE_DRIVEN_ATP_SYNTHESIS")
# Heatmap showing multiple rescued pathways
# Heatmap: brain region ~ pathway, value = PL vs AD NES
df.heatmap <- dcast(ID~brain_region, value.var="PLX5622_vs_FADCtrl", data=df.wide.sub.up_rescued)
df.heatmap <- data.frame(df.heatmap[, -1], row.names=df.heatmap$ID)
df.heatmap[is.na(df.heatmap)] <- 0
# Add representative genes for each pathway.
df.wide.sub.up_rescued.unique <- subset(df.wide.sub.up_rescued, !duplicated(ID))
mapping <- sapply(df.wide.sub.up_rescued.unique$genelist,
       function(x) {
         paste(strsplit(x, ",")[[1]][1:3], collapse=",")
         })
names(mapping) <- df.wide.sub.up_rescued.unique$ID
row.names(df.heatmap) <- paste0(row.names(df.heatmap), " | ", mapping[row.names(df.heatmap)])
library(pheatmap)
p <- pheatmap(df.heatmap)
```

#### Specific Pathway GSEAplot
```{r}
# GOCC_RESPIRATORY_CHAIN_COMPLEX
library(enrichplot)
pathway <- "GOCC_RESPIRATORY_CHAIN_COMPLEX"
BR <- "Thalamus"
CP <- "PLX5622_vs_FADCtrl"
out.path <- paste0("../001ST/001Giotto_out/003DEGs/GSEA/", BR, ".", CP, ".gsea.rds")
gsea_result <- readRDS(out.path)
gene.list <- subset(gsea_result@result, ID == pathway)$core_enrichment
p1 <- gseaplot2(gsea_result, pathway, title=paste0(pathway, "\nin ", BR, "\n", CP),
          pvalue_table = T, pvalue_table_rownames = NULL)
CP <- "FADCtrl_vs_WTCtrl"
out.path <- paste0("../001ST/001Giotto_out/003DEGs/GSEA/", BR, ".", CP, ".gsea.rds")
gsea_result <- readRDS(out.path)
p2 <- gseaplot2(gsea_result, pathway, title=paste0(pathway, "\nin ", BR, "\n", CP),
          pvalue_table = T, pvalue_table_rownames = NULL)
pathway <- "GOBP_CARBON_DIOXIDE_TRANSPORT"
BR <- "Olfactory area"
CP <- "PLX5622_vs_FADCtrl"
out.path <- paste0("../001ST/001Giotto_out/003DEGs/GSEA/", make.names(BR), ".", CP, ".gsea.rds")
gsea_result <- readRDS(out.path)
p1 <- gseaplot2(gsea_result, pathway, title=pathway,
          pvalue_table = T, pvalue_table_rownames = NULL)
CP <- "FADCtrl_vs_WTCtrl"
out.path <- paste0("../001ST/001Giotto_out/003DEGs/GSEA/", make.names(BR), ".", CP, ".gsea.rds")
gsea_result <- readRDS(out.path)
p2 <- gseaplot2(gsea_result, pathway, title=pathway,
          pvalue_table = T, pvalue_table_rownames = NULL)
gene.list <- subset(gsea_result@result, ID == pathway)$core_enrichment
```

#### Identify Down-regulated Rescued Pathways
```{r}
# Down-rescued and microglia-free pathways?
df.wide.sub.down_rescued <- subset(df.wide.sub, FADCtrl_vs_WTCtrl > 0 & PLX5622_vs_FADCtrl < 0)
df.wide.sub.down_rescued$genelist <- add_genelist(df.wide.sub.down_rescued)
nrow(df.wide.sub.down_rescued)
# [1] 1974
df.wide.sub.down_rescued <- subset(df.wide.sub.down_rescued, !grepl("C1qa|Trem2|Cd74|H2-Eb1|H2-Q7", df.wide.sub.down_rescued$genelist))
nrow(df.wide.sub.down_rescued)
# C1qa|Trem2|Cd74
# 1870,1322,1008,588,480
view(df.wide.sub.down_rescued)
# Brain region-specific down-rescued pathways?
df <- aggregate(brain_region~ID, FUN=length, data=df.wide.sub.down_rescued)
df <- subset(df, brain_region == 1)
df2 <- subset(df.wide.sub.down_rescued, ID %in% df$ID)
df2 <- df2[order(df2$brain_region), ]
# Down-rescued pathways containing brain region marker genes? Brain region-specific markers should be identified using WT.
br.markers <- read.csv("../001ST/001Giotto_out/001preprocess/brain_region_markers_lfc1.csv", row.names=1)
br.markers.gl <- unique(br.markers$feats)
df2 <- df2[df2.keep, ]
# Down-rescued pathways containing neuron-specific marker genes?
anno.L1.marker <- read.csv("../002SC/004Seurat_out/002cell_type_anno/anno.L1.FindAllMarkers.csv", row.names=1)
neuro.marker <- unique(subset(anno.L1.marker, cluster %in% c("GABA", "Glu") & avg_log2FC > 0.5)$gene)
if (F) {
  micro.marker <- unique(subset(anno.L1.marker, cluster == "Micro" & avg_log2FC > 0.5)$gene)
  df.wide.sub.down_rescued <- subset(df.wide.sub, FADCtrl_vs_WTCtrl > 0 & PLX5622_vs_FADCtrl < 0)
  df.wide.sub.down_rescued$genelist <- add_genelist(df.wide.sub.down_rescued)
 
  df.sub <- exclude_rows_contain_any_genes(df.wide.sub.down_rescued, micro.marker)
  nrow(df.sub) # 84
} else {
  micro.marker <- unique(subset(anno.L1.marker, cluster == "Micro")$gene)
  df.wide.sub.down_rescued <- subset(df.wide.sub, FADCtrl_vs_WTCtrl > 0 & PLX5622_vs_FADCtrl < 0)
  df.wide.sub.down_rescued$genelist <- add_genelist(df.wide.sub.down_rescued)
 
  df.sub <- exclude_rows_contain_any_genes(df.wide.sub.down_rescued, micro.marker)
  nrow(df.sub) # 14
}
if (F) {
  # Further filtering leaves nothing.
  df.sub <- keep_rows_contain_any_genes(df.sub, neuro.marker)
  df.sub <- add_intersect_col(df.sub, neuro.marker)
}
library(enrichplot)
pathway <- "GOMF_NEUROPEPTIDE_RECEPTOR_BINDING"
BR <- "Isocortex"
CP <- "FADCtrl_vs_WTCtrl"
out.path <- paste0("../001ST/001Giotto_out/003DEGs/GSEA/", BR, ".", CP, ".gsea.rds")
gsea_result <- readRDS(out.path)
p2 <- gseaplot2(gsea_result, pathway, title=pathway,
          pvalue_table = T, pvalue_table_rownames = NULL)
CP <- "PLX5622_vs_FADCtrl"
out.path <- paste0("../001ST/001Giotto_out/003DEGs/GSEA/", BR, ".", CP, ".gsea.rds")
gsea_result <- readRDS(out.path)
gene.list <- subset(gsea_result@result, ID == pathway)$core_enrichment
p1 <- gseaplot2(gsea_result, pathway, title=pathway,
          pvalue_table = T, pvalue_table_rownames = NULL)
pathway <- "GOMF_HYDROLASE_ACTIVITY_HYDROLYZING_N_GLYCOSYL_COMPOUNDS"
BR <- "Olfactory area"
CP <- "FADCtrl_vs_WTCtrl"
out.path <- paste0("../001ST/001Giotto_out/003DEGs/GSEA/", make.names(BR), ".", CP, ".gsea.rds")
gsea_result <- readRDS(out.path)
p1 <- gseaplot2(gsea_result, pathway, title=pathway,
          pvalue_table = T, pvalue_table_rownames = NULL)
CP <- "PLX5622_vs_FADCtrl"
out.path <- paste0("../001ST/001Giotto_out/003DEGs/GSEA/", make.names(BR), ".", CP, ".gsea.rds")
gsea_result <- readRDS(out.path)
gene.list <- subset(gsea_result@result, ID == pathway)$core_enrichment
p2 <- gseaplot2(gsea_result, pathway, title=pathway,
          pvalue_table = T, pvalue_table_rownames = NULL)
pathway <- "GOMF_CHEMOKINE_ACTIVITY"
BR <- "Isocortex"
CP <- "FADCtrl_vs_WTCtrl"
out.path <- paste0("../001ST/001Giotto_out/003DEGs/GSEA/", make.names(BR), ".", CP, ".gsea.rds")
gsea_result <- readRDS(out.path)
p1 <- gseaplot2(gsea_result, pathway, title=pathway,
          pvalue_table = T, pvalue_table_rownames = NULL)
CP <- "PLX5622_vs_FADCtrl"
out.path <- paste0("../001ST/001Giotto_out/003DEGs/GSEA/", make.names(BR), ".", CP, ".gsea.rds")
gsea_result <- readRDS(out.path)
gene.list <- subset(gsea_result@result, ID == pathway)$core_enrichment
p2 <- gseaplot2(gsea_result, pathway, title=pathway,
          pvalue_table = T, pvalue_table_rownames = NULL)
pathway <- "GOMF_MHC_PROTEIN_BINDING"
BR <- "Isocortex"
CP <- "FADCtrl_vs_WTCtrl"
out.path <- paste0("../001ST/001Giotto_out/003DEGs/GSEA/", make.names(BR), ".", CP, ".gsea.rds")
gsea_result <- readRDS(out.path)
p1 <- gseaplot2(gsea_result, pathway, title=pathway,
          pvalue_table = T, pvalue_table_rownames = NULL)
CP <- "PLX5622_vs_FADCtrl"
out.path <- paste0("../001ST/001Giotto_out/003DEGs/GSEA/", make.names(BR), ".", CP, ".gsea.rds")
gsea_result <- readRDS(out.path)
gene.list <- subset(gsea_result@result, ID == pathway)$core_enrichment
p2 <- gseaplot2(gsea_result, pathway, title=pathway,
          pvalue_table = T, pvalue_table_rownames = NULL)
```

#### Mitochondrial Gene Investigation: PL12 and PL14
What are the pathways with the highest overlap with PL12 and PL14, and what are the trends of changes between groups?
```{r}
df.merged <- read.csv("../001ST/001Giotto_out/003DEGs/GSEA/st.gsea.csv", row.names=1)
df.merged.sub <- subset(df.merged, grepl("mt-Nd1", df.merged$core_enrichment))
df.merged.sub <- subset(df.merged, ID=="GOBP_OXIDATIVE_PHOSPHORYLATION", select=c("brain_region", "comparison", "NES", "ID"))
# Selection criterion 1: Pathways with the highest proportion of mt genes
# Selection criterion 2: Pathways containing Ndufb and other PL12 genes with the fewest mt genes.
# Compare the inter-group change patterns of the two types of pathways.
df.merged.sub <- keep_rows_contain_any_genes(df.merged, "mt-Nd3", gl.col="core_enrichment", gl.sep="/")
df.merged.sub <- exclude_rows_contain_any_genes(df.merged.sub, "Ndufb7")
# Pathways with the highest enrichment score with PL12.
# Pathways with the highest enrichment score with PL14.
# Compare the inter-group change patterns of the two types of pathways.
# Help me write a function in R to calculate the Jaccard coefficient of two string vectors
jaccard_strings <- function(vec1, vec2) {
  # Convert to unique sets (deduplicate)
  set1 <- unique(vec1)
  set2 <- unique(vec2)
 
  # Calculate the size of intersection and union
  intersection <- length(intersect(set1, set2))
  union_size <- length(union(set1, set2))
 
  # Jaccard coefficient = intersection / union
  if (union_size == 0) return(1) # Two empty sets are defined as 1 (can be changed to 0 as needed)
 
  intersection / union_size
}
mod.df.list <- readRDS("../001ST/001Giotto_out/007spatial_coexpr/hdWGCNA/wgcna_modules.list.rds")
PL.df <- mod.df.list$PLX5622
PL12.gene <- subset(PL.df, module == "PL12")$gene_name
PL14.gene <- subset(PL.df, module == "PL14")$gene_name
df.merged$PL12.jaccard <- sapply(df.merged$core_enrichment, function(x) {
  # y <- strsplit(df.merged.sub$core_enrichment[1], "/")[[1]]
  y <- strsplit(x, "/")[[1]]
  jaccard_strings(y, PL12.gene)
})
df.merged$PL14.jaccard <- sapply(df.merged$core_enrichment, function(x) {
  # y <- strsplit(df.merged.sub$core_enrichment[1], "/")[[1]]
  y <- strsplit(x, "/")[[1]]
  jaccard_strings(y, PL14.gene)
})
# Is PL14 upregulated in AD vs WT?
df.merged <- df.merged[order(df.merged$PL14.jaccard, decreasing = T), ]
df.merged.sub <- subset(df.merged, comparison == "FADCtrl_vs_WTCtrl" & PL14.jaccard > 0.2)
View(df.merged.sub)
# Answer: Basically downregulated. Only in Fiber tracts.
df.merged.sub <- subset(df.merged, comparison == "PLX5622_vs_FADCtrl" & PL14.jaccard > 0.2)
View(df.merged.sub)
unique(df.merged.sub$brain_region)
# [1] "Olfactory area" "Striatum" "Meninges" "Thalamus"
# In PL vs AD, all are upregulated, but cover more brain regions.
# Now check PL12
df.merged <- df.merged[order(df.merged$PL12.jaccard, decreasing = T), ]
df.merged.sub <- subset(df.merged, comparison == "FADCtrl_vs_WTCtrl" & PL14.jaccard < 0.1 & PL12.jaccard > 0.1)
View(df.merged.sub)
unique(df.merged.sub$brain_region)
# [1] "Isocortex" "Fiber tracts" "Hippocampus"
# All downregulated in AD vs WT, but compared to PL14, it includes more brain regions (hippocampus, cortex).
df.merged.sub <- subset(df.merged, comparison == "PLX5622_vs_FADCtrl" & PL14.jaccard < 0.1 & PL12.jaccard > 0.1)
View(df.merged.sub)
unique(df.merged.sub$brain_region)
# "Striatum" "Thalamus"
# All upregulated, mainly including striatum and thalamus.
#------------------------------------------------------------
# Plotting: Select representative pathways and draw heatmaps for the two comparison groups separately.
# 1. Custom color gradient: low (blue) → 0 (white) → high (red)
my_colors <- colorRampPalette(c("navy", "white", "firebrick3"))(100)
# 2. Or more explicitly control the position of 0 (recommended)
my_breaks <- c(seq(-3, 0, length.out = 50), # Negative values to 0 → blue to white
               seq(0.0001, 3, length.out = 50)) # Positive values → white to red
# PL14
PL14.pathway <- unique(subset(df.merged, PL14.jaccard > 0.3)$ID)
df.merged.sub <- subset(df.merged, ID %in% PL14.pathway,
                        select=c("brain_region", "comparison", "ID", "NES"))
df.merged.sub$brain_region.uniq <- paste0(df.merged.sub$brain_region, ifelse(df.merged.sub$comparison == "FADCtrl_vs_WTCtrl", "1", "2"))
df.heatmap <- dcast(ID~brain_region.uniq, value.var="NES", data=df.merged.sub)
df.heatmap <- data.frame(df.heatmap[, -1], row.names=df.heatmap$ID)
df.heatmap[is.na(df.heatmap)] <- 0
pheatmap(df.heatmap, color = my_colors, breaks = my_breaks, cluster_cols = F)
# PL12
PL12.pathway <- unique(subset(df.merged, PL14.jaccard < 0.1 & PL12.jaccard > 0.1)$ID)
df.merged.sub <- subset(df.merged, ID %in% PL12.pathway,
                        select=c("brain_region", "comparison", "ID", "NES"))
df.merged.sub$brain_region.uniq <- paste0(df.merged.sub$brain_region, ifelse(df.merged.sub$comparison == "FADCtrl_vs_WTCtrl", "1", "2"))
df.heatmap <- dcast(ID~brain_region.uniq, value.var="NES", data=df.merged.sub)
df.heatmap <- data.frame(df.heatmap[, -1], row.names=df.heatmap$ID)
df.heatmap[is.na(df.heatmap)] <- 0
vec <- colnames(df.heatmap)
df.heatmap <- df.heatmap[, vec[order(substr(vec, nchar(vec), nchar(vec)))]]
pheatmap(df.heatmap, color = my_colors, breaks = my_breaks, cluster_cols = F)
```

### Retrieve Pathways Containing Specific Genes
```{r}
df.merged[grepl("Bnc2", df.merged$core_enrichment), ]
view(df.merged[grepl("Nrp1", df.merged$core_enrichment), ])
gene.list <- c("Gm41609", "Bnc2", "Mical2", "9630028H03Rik", "Plk2", "Tfap2b",
                                                 "Rorb", "Plk5", "Pax5", "Rmst", "Gria1")
get_rows_contain_genes_in_df.merged(df.merged, gene.list)
view(df %>% filter(brain_region=="Hypothalamus") %>% arrange(qvalue))
view(get_rows_contain_genes_in_df.merged(df.merged, c("Hcrt")))
```

### gseaplot
```{r}
library(enrichplot)
pathway <- "GOBP_HYPOTHALAMUS_CELL_DIFFERENTIATION"
out.path <- "../001ST/001Giotto_out/003DEGs/GSEA/Hypothalamus.PLX5622_vs_FADCtrl.gsea.rds"
gsea_result <- readRDS(out.path)
p1 <- gseaplot2(gsea_result, pathway, title=pathway,
          pvalue_table = T, pvalue_table_rownames = NULL)
# Nrp1/Otp/Ndnf/Nrp2/Nhlh2/Pou3f2
if (F) {
  # Not significant
  out.path <- "../001ST/001Giotto_out/003DEGs/GSEA/Hypothalamus.FADCtrl_vs_WTCtrl.gsea.rds"
  gsea_result <- readRDS(out.path)
  p2 <- gseaplot2(gsea_result, pathway, title=pathway,
            pvalue_table = T, pvalue_table_rownames = NULL)
}
# pathway <- "GOMF_MHC_PROTEIN_BINDING"
BR <- "Hypothalamus"
CP <- "PLX5622_vs_FADCtrl"
out.path <- paste0("../001ST/001Giotto_out/003DEGs/GSEA/", make.names(BR), ".", CP, ".gsea.rds")
gsea_result <- readRDS(out.path)
p1 <- gseaplot2(gsea_result, pathway, title=pathway,
          pvalue_table = T, pvalue_table_rownames = NULL)
# GOBP_WNT_SIGNALING_PATHWAY
# GOBP_REGULATION_OF_WNT_SIGNALING_PATHWAY
# GOBP_REGULATION_OF_WNT_SIGNALING_PATHWAY
```
