
---
output: html_document
editor_options:
  chunk_output_type: console
---

# Spot DEGs Pathway Enrichment Analysis (new)
## Metascape Enrichment Analysis
Compared with the old version, this version of the enrichment analysis has the following improvements:

1. Background was added. When performing Metascape enrichment analysis, genes expressed in the spatial transcriptomics data were used as the background.
2. No log2FC threshold is set. Any gene with padj < 0.05 is considered a DEG.

The following code was migrated from `figure_process.Rmd`. Please update the referenced paths accordingly.

```{r}
# fig. 4h: ST DEGs pathway enrichment heatmap ----
library(dplyr)
# 1. Obtain gene lists (excluding ventricles) and organize into multiple gene lists.
st.DEGs <- read.csv("data/st.DEGs.csv", row.names = 1)
st.DEGs.sub <- subset(st.DEGs, comparison %in% c("FADCtrl_vs_WTCtrl", "PLX5622_vs_FADCtrl"))
st.DEGs.sub$direction <- ifelse(st.DEGs.sub$avg_log2FC > 0, "up", "down")
st.DEGs.agg <- st.DEGs.sub %>% group_by(comparison, brain_region, direction) %>%
  summarise(
    gene_list = paste(`gene`, collapse = ","), # Method 1: comma-separated concatenation
    gene_count = n(), # Method 2: count
    .groups = "drop" # Automatically ungroup after summarization
  )
write.csv(st.DEGs.agg, "../data/fig.4h_st.DEGs.agg.csv", row.names=F)
# 2. Perform Metascape pathway enrichment analysis. Since a maximum of 30 gene lists are allowed per analysis,
# the analysis was previously split into three comparison groups. Now we only focus on PL vs AD and AD vs WT,
# so we can split into only two comparison groups.
st.DEGs.agg$list_name <- paste0(st.DEGs.agg$brain_region, "__", st.DEGs.agg$direction)
st.DEGs.agg <- subset(st.DEGs.agg, ! brain_region %in% c("Lateral Ventricle", "Third Ventricle"))
df1 <- subset(st.DEGs.agg, comparison == "FADCtrl_vs_WTCtrl", select=c("list_name", "gene_list"))
df2 <- subset(st.DEGs.agg, comparison == "PLX5622_vs_FADCtrl", select=c("list_name", "gene_list"))
write.table(df1, "../data/st.DEGs.agg_AD_vs_WT.txt", quote = F, sep = "\t", col.names = F, row.names = F)
write.table(df2, "../data/st.DEGs.agg_PL_vs_AD.txt", quote = F, sep = "\t", col.names = F, row.names = F)
# Prepare background file
# pseudocount
# How to operate on a sparse matrix (dgCMatrix)?
# 1. How to set values greater than 0 to 1.
# 2. Aggregate by column. You can use functions from scuttle.
gobj <- ""
# gobj.expr <- gobj@expression$cell$rna$raw
# spot.meta_info <- pDataDT(gobj)[, c("cell_types", "list_ID")]
# spot.meta_info.list <- paste0(spot.meta_info$cell_types, ".", spot.meta_info$list_ID)
#
# expressed_spots <- lapply(1:20, function(i){
# gobj.expr[i, ] != 0
# df <- table(spot.meta_info.list[gobj.expr[i, ] != 0])
# df
# })
#
# do.call(cbind, expressed_spots)
library(SingleCellExperiment)
meta <- data.frame(gobj@cell_metadata$cell$rna@metaDT)
row.names(meta) <- meta$cell_ID
gobj.count <- CreateSeuratObject(gobj@expression$cell$rna$raw@exprMat > 0,
                                 meta.data=meta)
sce <- as.SingleCellExperiment(gobj.count)
summed <- scuttle::aggregateAcrossCells(sce,
                                        id=colData(sce)[,c("cell_types", "list_ID")])
head(assays(summed)$counts)
# #
# gobj.count2 <- CreateSeuratObject(gobj@expression$cell$rna$raw@exprMat,
# meta.data=meta)
# sce2 <- as.SingleCellExperiment(gobj.count2)
# summed2 <- scuttle::aggregateAcrossCells(sce2,
# id=colData(sce2)[,c("cell_types", "list_ID")])
# Filter
df <- assays(summed)$counts
nrow(assays(summed)$counts)
# [1] 18551
keep <- apply(df, 1, function(x){any(x >= 10)})
df.filtered <- df[keep, ]
nrow(df.filtered)
# [1] 16006
bg.gene.list <- row.names(df.filtered)
writeLines(bg.gene.list, "../data/st.metascape.bg.gene.list.txt")
# 3. Organize the enrichment analysis results into tables for teachers to screen.
# 4. Draw heatmaps.
# Directly used the heatmap generated by Metascape.
# 5. Count gene lists
rescued_genes <- ""
PL_vs_AD.metascape <- read.csv("../data/st.DEGs.agg_PL_vs_AD.txt_custom_bg/Enrichment_GO/GO_AllLists.csv")
# Key columns: GeneList, GO
PL_vs_AD.df <- data.frame(GO=PL_vs_AD.metascape$GO,
                          brain_region = sapply(strsplit(PL_vs_AD.metascape$GeneList, "__"), "[", 1),
                          direction.PL = sapply(strsplit(PL_vs_AD.metascape$GeneList, "__"), "[", 2))
AD_vs_WT.metascape <- read.csv("../data/st.DEGs.agg_AD_vs_WT.txt_custom_bg/Enrichment_GO/GO_AllLists.csv")
AD_vs_WT.df <- data.frame(GO=AD_vs_WT.metascape$GO,
                          brain_region = sapply(strsplit(AD_vs_WT.metascape$GeneList, "__"), "[", 1),
                          direction.AD = sapply(strsplit(AD_vs_WT.metascape$GeneList, "__"), "[", 2))
df.merged <- merge(PL_vs_AD.df, AD_vs_WT.df, by=c("GO", "brain_region"))
df.merged$pattern <- paste0(df.merged$direction.AD, "_", df.merged$direction.PL)
table(df.merged$pattern)
# down_down down_up up_down up_up
# 182 620 1950 122
up_rescued_pathways <- subset(df.merged, pattern == "down_up")
down_rescued_pathways <- subset(df.merged, pattern == "up_down")
up_rescued_pathways <- unique(up_rescued_pathways$GO)
length(up_rescued_pathways)
# 153
down_rescued_pathways <- unique(down_rescued_pathways$GO)
length(down_rescued_pathways)
# 687
pathway.heatmap.PL <- read.csv("../data/st.DEGs.agg_PL_vs_AD.txt_custom_bg/Enrichment_heatmap/HeatmapSelectedGO.csv")
pathway.heatmap.PL.up <- pathway.heatmap.PL[pathway.heatmap.PL$GO %in% up_rescued_pathways, ]
pathway.heatmap.PL.down <- pathway.heatmap.PL[pathway.heatmap.PL$GO %in% down_rescued_pathways, ]
```

## Glycolysis
How many genes are there?
1. Read the enrichment analysis result file `GO_AllLists.csv` for PL vs AD.
2. Subset the glycolysis pathway, extract the genes, and count them.

```{r}
spot.DEGs <- read.csv("../001ST/001Giotto_out/003DEGs/wilcox/st.DEGs.csv", row.names=1)
df.enrich <- read.csv("../001ST/001Giotto_out/003DEGs/wilcox/enrich2/st.DEGs.agg_PL_vs_AD.txt_custom_bg/Enrichment_GO/GO_AllLists.csv")
df.sub <- subset(df.enrich, GO == "WP157")
gl <- unique(unlist(strsplit(df.sub$Hits, "|", fixed = T)))
gl.n <- length(gl)
gl.collapsed <- paste(gl, collapse=", ")
print(paste0("In PL vs AD, a total of ", gl.n, " genes were enriched in the Glycolysis and gluconeogenesis pathway. These genes include: ", gl.collapsed, ". "))
# [1] "In PL vs AD, a total of 14 genes were enriched in the Glycolysis and gluconeogenesis pathway. These genes include: Eno2, Ldha, Mdh2, Pkm, Tpi1, Gapdh, Got1, Got2, Gpi1, Mdh1, Pfkl, Pgam1, Pgk1, Pfkp. "
subset(spot.DEGs, gene %in% gl & comparison == "PLX5622_vs_FADCtrl")
```

## Oxidative Phosphorylation
How many genes? Which genes?

```{r number-of-oxphos-genes}
for (ID in c("mmu00190", "GO:0006119")) {
 
  df.sub <- subset(df.enrich, GO == ID) # Oxidative phosphorylation
  description <- df.sub[1, "Description"]
 
  gl <- unique(unlist(strsplit(df.sub$Hits, "|", fixed = T)))
 
  gl.n <- length(gl)
 
  gl.collapsed <- paste(gl, collapse=", ")
 
  print(paste0("In PL vs AD, a total of ", gl.n, " genes were enriched in the ",
               description, " pathway. These genes include: ",
               gl.collapsed, ". "))
}
mapping <- c("ATP6"="mt-Atp6", "ATP8"="mt-Atp8",
             "COX1"="mt-Co1", "COX2"="mt-Co2", "COX3"="mt-Co3",
             "ND1"="mt-Nd1", "ND2"="mt-Nd2", "ND3"="mt-Nd3",
             "ND4"="mt-Nd4", "ND4L"="mt-Nd4l", "ND5"="mt-Nd5", "ND6"="mt-Nd6", "CYTB"="mt-Cytb")
df.sub <- subset(df.enrich, GO == "GO:0006119")
gl <- unique(unlist(strsplit(df.sub$Hits, "|", fixed = T)))
gl2 <- c()
for (i in gl) {
  if (i %in% names(mapping)) {
    i <- mapping[i]
  }
  gl2 <- c(gl2, i)
}
# Genes not in ST DEGs:
gl3 <- gl2[!gl2 %in% spot.DEGs$gene]
gl3
# "Atp5me" "Ndufb1" "Atp5pd"
# Question 1: Why are these genes not in st.DEGs but were captured by the enrichment analysis?
# Check the st.DEGs.agg_PL_vs_AD.txt file and find that these genes are not present at all. It may still be an issue of gene name aliases.
# Atp5me = Atp5k
# Ndufb1 = Ndufb1-ps
# Atp5pd = Atp5h
# Question 2: Why does it say 30 genes enriched in oxidative phosphorylation, but only 23 are listed here?
# Notably, 23 of them were associated with oxidative phosphorylation (OXPHOS) pathway (GO:0006119) and formed a densely connected module (Fig. 4 Ext. d; Supplementary Table 4), including Ndufa3, Ndufa5, mt-Co2, Cox7c, mt-Nd1, mt-Nd2, mt-Nd3, mt-Nd4, mt-Nd4l, and mt-Nd5 (Fig. 4 Ext. e).
# Answer: Because it refers to rescued genes here.
# Question 3: Why does it say 30 genes enriched in oxidative phosphorylation, but only 17 upregulated genes are listed here?
# Our ST profiling confirmed downregulation of the OXPHOS pathway in AD mice and revealed that microglial depletion at the mid-disease stage upregulated expression of 17 OXPHOS genes in AD mouse brain (Fig. 4 Ext. d; Fig. 6a).
# Answer: It seems I only included genes from two clusters. It should be:
# 1. List all upregulated oxidative phosphorylation genes (30? based on MitoCarta or MGI database).
# 2. Then list the up-rescued ones (23?).
# 3. Then list the ones belonging to the electron transport chain within the up-rescued set (13?).
# Question 4: Why does Fig. 6b only show 13 genes?
# Answer: Because only rescued genes were included.
```

## Translation
GO TERM: GO:0006412
```{r}
spot.DEGs <- read.csv("../001ST/001Giotto_out/003DEGs/wilcox/st.DEGs.csv", row.names=1)
df.enrich <- read.csv("../001ST/001Giotto_out/003DEGs/wilcox/enrich2/st.DEGs.agg_PL_vs_AD.txt_custom_bg/Enrichment_GO/GO_AllLists.csv")
df.sub <- subset(df.enrich, GO == "GO:0006412")
gl <- unique(unlist(strsplit(df.sub$Hits, "|", fixed = T)))
gl.n <- length(gl)
gl.collapsed <- paste(gl, collapse=", ")
print(paste0("In PL vs AD, a total of ", gl.n, " genes were enriched in the translation pathway. These genes include: ", gl.collapsed, ". "))
```

```
[1] "In PL vs AD, a total of 54 genes were enriched in the translation pathway. These genes include: Rpl37a, Rps19, Rps29, Rps28, Rps27, Rps23, Rps21, Rpl39, Rpl37, Rpl38, Rpl41, Eef1a1, Fau, Hspa5, Rpl19, Rpl21, Rpl32, Rpl9, Uba52, Rpl27a, Eif2s3y, Rpl11, Rpl22l1, Rps15a, Rps25, Rpl5, Rpl35, Rpl4, Rpl31, Rpl26, Rps14, Rps18, Rps8, Rps26, Rpl35a, Rpl23, Rplp2, Lars2, Rpl23a, Rpl17, Eif4a2, Rheb, Rpl10a, Rpl22, Rpl29, Rpl30, Rps12, Rps4x, Xbp1, Rplp1, Rps20, Rpl18a, Plxnb2, Rpl13. "
```

# Spot DEGs GSEA
## Preliminary Investigation
I now want to perform GSEA analysis using the results from `Seurat::FindMarkers`. After obtaining the differential expression gene table, should I filter based on `logfc.threshold` or basal expression level?

Gemini suggests not to filter, and to include as many genes as possible. Please refer to the appendix: Preprocessing for GSEA.

`Seurat::FindMarkers` applies filtering by default with the following thresholds:  
`logfc.threshold = 0.25`, `min.pct = 0.1`

It may be necessary to disable filtering to obtain a more complete log2FC vector for GSEA analysis.

Questions:
- Should a filtering threshold be set (e.g., basal expression level)?
- Should wilcox or MAST be used? Previously wilcox was used. Switching to MAST would require updating a lot of data. However, MAST results appear more reliable because they account for batch effects.

Input:
- testcombo.rds, Giotto object  
Output:
- DEGs table for each brain region  

Process:
- Read the Giotto object and convert it to a Seurat object.
- Use `Seurat::FindMarkers` to identify differentially expressed genes.

## Load Data
```{r}
library(Giotto)
library(Seurat)
gobj <- readRDS("001Giotto_out/testcombo.rds")
Giotto2Seurat <- function(obj){
  expr <- get_expression_values(obj)@exprMat
  meta <- as.data.frame(pDataDT(obj))
  row.names(meta) <- meta$cell_ID
  obj.seurat <- CreateSeuratObject(counts=expr, meta.data=meta)
  obj.seurat <- NormalizeData(obj.seurat)
  return(obj.seurat)
}
sobj <- Giotto2Seurat(gobj)
sobj$treatment__br <- paste0(sobj$treatment, "__", sobj$cell_types)
Idents(sobj) <- "treatment__br"
str(sobj@meta.data)
```

## Identify Spot DEGs
During execution, MAST reported an error. The solution can be found in the appendix: MAST issues.

```{r}
CP <- "PLX5622_vs_FADCtrl"
TR1 <- strsplit(CP, "_vs_")[[1]][1]
TR2 <- strsplit(CP, "_vs_")[[1]][2]
BR <- "Olfactory area" # Brain region
TR_BR1 <- paste0(TR1, "__", BR)
TR_BR2 <- paste0(TR2, "__", BR)
df.deg.1 <- FindMarkers(sobj, ident.1=TR_BR1, ident.2=TR_BR2, test.use="wilcox")
df.deg.2 <- FindMarkers(sobj, ident.1=TR_BR1, ident.2=TR_BR2, test.use="wilcox", logfc.threshold=0)
df.deg.3 <- FindMarkers(sobj, ident.1=TR_BR1, ident.2=TR_BR2, test.use="MAST", latent.vars="batch")
df.deg.4 <- FindMarkers(sobj, ident.1=TR_BR1, ident.2=TR_BR2, test.use="MAST", latent.vars="batch", logfc.threshold=0)
nrow(df.deg.1) # 136
nrow(df.deg.2) # 10817
length(intersect(row.names(df.deg.3), row.names(df.deg.1))) # 136
```

```{r}
dir.create("001Giotto_out/003DEGs/wilcox/GSEA")
for (CP in c("PLX5622_vs_FADCtrl", "FADCtrl_vs_WTCtrl")) {
  TR1 <- strsplit(CP, "_vs_")[[1]][1]
  TR2 <- strsplit(CP, "_vs_")[[1]][2]
  for (BR in setdiff(unique(sobj$cell_types), c("C12", "Third Ventricle", "Lateral Ventricle"))) {
    TR_BR1 <- paste0(TR1, "__", BR)
    TR_BR2 <- paste0(TR2, "__", BR)
   
    cat("current run: ", BR, CP, "\n")
   
    df.gsea <- FindMarkers(
      sobj,
      ident.1 = TR_BR1,
      ident.2 = TR_BR2,
      test.use = "wilcox",
     
      # Key parameter settings:
      logfc.threshold = 0, # Do not filter logFC, we need all genes
      min.pct = 0.01, # Keep a minimal threshold to filter out noisy genes that are barely expressed (e.g., only in 1-2 cells). This speeds up computation without affecting GSEA.
      return.thresh = 1 # Return all p-values, do not return only significant results
    )
    out.path <- paste0("001Giotto_out/003DEGs/wilcox/GSEA/", BR, ".", CP, ".csv")
    write.csv(df.gsea, out.path)
  }
}
nrow(df.gsea)
# [1] 15267
```

## Preprocess Data
```{r}
library(dplyr)
library(tibble)
# Preprocess data
prepare_gene_list <- function(df.gsea) {
  gsea.rank <- df.gsea %>%
    rownames_to_column("gene") %>%
    # 1. Handle cases where p_val == 0 (to avoid Inf from -log10(0))
    # Replace 0 with a value slightly smaller than the smallest non-zero p-value in the data
    mutate(p_val = ifelse(p_val == 0, .Machine$double.xmin, p_val)) %>%
   
    # 2. Calculate ranking metric
    mutate(rank_metric = sign(avg_log2FC) * -log10(p_val)) %>%
   
    # 3. Sort (from largest to smallest)
    arrange(desc(rank_metric)) %>%
   
    # 4. Extract pure vector for clusterProfiler
    select(gene, rank_metric)
 
  # Convert to named vector (format required by GSEA function)
  gene_list <- gsea.rank$rank_metric
  names(gene_list) <- gsea.rank$gene
  return(gene_list)
}
# # Check
# head(gene_list)
# # Should be positive values representing highly significant upregulation
# tail(gene_list)
# # Should be negative values representing highly significant downregulation
```

## Run GSEA
```{r}
# env: thinkpad
library(clusterProfiler)
library(msigdbr) # Obtain gene sets
Brain_regions <- strsplit("Isocortex, Hypothalamus, Meninges, Fiber tracts, Cortical subplate, Thalamus, Olfactory area, Striatum, Hippocampus", ", ")[[1]]
for (CP in c("PLX5622_vs_FADCtrl", "FADCtrl_vs_WTCtrl")) {
  for (BR in Brain_regions) {
    in.path <- paste0("../001ST/001Giotto_out/003DEGs/wilcox/GSEA/", BR, ".", CP, ".csv")
    df.gsea <- read.csv(in.path, row.names=1)
   
    gene_list <- prepare_gene_list(df.gsea)
    COLLECT <- "C5"
   
    # Assume you are working with mouse data, obtain the C5 gene set
    m_t2g <- msigdbr(species = "Mus musculus", collection = COLLECT) %>%
      dplyr::select(gs_name, gene_symbol)
   
    res.gsea <- GSEA(gene_list,
                     TERM2GENE = m_t2g,
                     pvalueCutoff = 0.1,
                     seed = 123) # Set seed for reproducibility
   
    out.path <- paste0("../001ST/001Giotto_out/003DEGs/wilcox/GSEA/", BR, ".", CP, ".gsea.rds")
    saveRDS(res.gsea, out.path)
  }
}
```

## GSEA Plot
Oxidative phosphorylation pathway in Olfactory area for both AD vs WT and PL vs AD.

```{r}
library(enrichplot)
pathway <- "GOBP_OXIDATIVE_PHOSPHORYLATION"
BR <- "Olfactory area"
CP <- "PLX5622_vs_FADCtrl"
in.path <- paste0("../001ST/001Giotto_out/003DEGs/wilcox/GSEA/", BR, ".", CP, ".gsea.rds")
gsea_result <- readRDS(in.path)
p1 <- gseaplot(gsea_result, geneSetID = pathway, by = "runningScore")
p1@layers$geom_linerange <- NULL
p1@layers$geom_line <- NULL
pvalue <- round(subset(gsea_result@result, ID==pathway)$p.adjust, 3)
title <- "GO: Oxidative phosphorylation\n5xFAD-PLX vs 5xFAD-CTRL"
p1 <- p1 + ggtitle(title,
                   subtitle = "Adjusted <i>P</i> < 0.0001") +
  geom_linerange(aes(x = x, ymin = ymin, ymax = ymax),
                 linewidth = 0.1, # or use size = 1.5 in older ggplot2
                 color = "black") + 
  geom_line(aes(y=runningScore), color="green", size=0.5) +
  theme_nature() + theme(plot.subtitle = element_markdown(size=8))
CP <- "FADCtrl_vs_WTCtrl"
in.path <- paste0("../001ST/001Giotto_out/003DEGs/wilcox/GSEA/", BR, ".", CP, ".gsea.rds")
gsea_result <- readRDS(in.path)
p2 <- gseaplot(gsea_result, geneSetID = pathway, by = "runningScore")
p2@layers$geom_linerange <- NULL
p2@layers$geom_line <- NULL
pvalue <- round(subset(gsea_result@result, ID==pathway)$p.adjust, 3)
title <- "GO: Oxidative phosphorylation\n5xFAD-CTRL vs WT-CTRL"
p2 <- p2 + ggtitle(title,
                   subtitle = "Adjusted <i>P</i> < 0.0001") +
  geom_linerange(aes(x = x, ymin = ymin, ymax = ymax),
                 linewidth = 0.1, 
                 color = "black") + 
  geom_line(aes(y=runningScore), color="green", size=0.5) +
  theme_nature() + theme(plot.subtitle = element_markdown(size=8))
p <- p2 | p1
width = 4.97
height=2.41
ggsave(paste0("../001ST/001Giotto_out/003DEGs/wilcox/GSEA/", pathway, ".", CP, ".", BR, ".png"),
       width=width, height=height, dpi=300)
ggsave(paste0("../001ST/001Giotto_out/003DEGs/wilcox/GSEA/", pathway, ".", CP, ".", BR, ".pdf"),
       width=width, height=height, device = cairo_pdf)
```
