
# prepare environment

scvelo is installed in conda environment named scvelo. 

Although velocyto was successfully installed in the `scvelo` environment, it kept reporting missing dependencies when used. It was better to just install it using the system's Python installation.

```{bash}

conda deactivate

sudo apt install python3.10-venv

python3 -m venv velocyto

source velocyto/bin/activate

pip install numpy scipy cython numba matplotlib scikit-learn h5py click

pip install pysam

git clone https://github.com/velocyto-team/velocyto.py.git

cd velocyto.py

pip install --no-build-isolation .

velocyto --help

<!-- Usage: velocyto [OPTIONS] COMMAND [ARGS]... -->

<!-- Options: -->
<!--   --version  Show the version and exit. -->
<!--   --help     Show this message and exit. -->

<!-- Commands: -->
<!--   run            Runs the velocity analysis outputting a loom file -->
<!--   run10x         Runs the velocity analysis for a Chromium Sample -->
<!--   run-dropest    Runs the velocity analysis on DropEst preprocessed data -->
<!--   run-smartseq2  Runs the velocity analysis on SmartSeq2 data (independent bam file per cell) -->
<!--   tools          helper tools for velocyto -->

# When actually running Velocyto, it will report an error saying that samtools is missing, so it's best to install it.
sudo apt install samtools

```

# run velocyto

```{bash}

# batch 1
for i in AD_ctrl WT_ctrl; do 
  mkdir -p cellranger.out/${i}.1/outs
  ln -s /storage/data/小胶质细胞与阿尔茨海默病/博奥经典单细胞数据-原始数据备份/KYSY-7672-221034-01/bam/${i}/* ./cellranger.out/${i}.1/outs
done

# batch 2
for i in AD_5622 AD_ctrl WT_ctrl; do 
  mkdir -p cellranger.out/${i}.2/outs
  ln -s /storage/data/小胶质细胞与阿尔茨海默病/博奥经典单细胞数据-原始数据备份/KYSY-8399-221034-02/bam/${i}/* ./cellranger.out/${i}.2/outs
done

# batch 3没有给bam file，先不管了。

# batch 4
# 数据在E:\project_backup\2022-10-14_mouse_spatial_v0.1\002SC\000data\batch4，得先拷贝到optiplex

for i in AD_5622 AD_ctrl WT_ctrl; do 
  mkdir -p cellranger.out/${i}.4/outs
  ln -s /storage/data/小胶质细胞与阿尔茨海默病/batch4/bam/${i}/* ./cellranger.out/${i}.4/outs
done


# 把filtered_feature_bc_matrix复制到./cellranger.out/AD_5622/outs下。

conda deactivate

source velocyto/bin/activate

# 已完成：AD_5622.1, AD_ctrl.1 WT_ctrl.1 AD_5622.2 AD_ctrl.2 WT_ctrl.2

for i in AD_5622.4 AD_ctrl.4 WT_ctrl.4; do 
  cp -r ./cellranger.out/${i} /scratch/project/
  velocyto run10x \
    /scratch/project/${i} \
    refdata-gex-mm10-2020-A/genes/genes.gtf
  rm /scratch/project/${i}/outs/cellsorted_possorted_genome_bam.bam
done

for i in AD_ctrl.1 WT_ctrl.1 AD_5622.2 AD_ctrl.2 WT_ctrl.2 AD_5622.4 AD_ctrl.4 WT_ctrl.4; do 
  cp -r /scratch/project/${i}/velocyto/ ./cellranger.out/${i}/
done
```

# install scvelo

```{bash}

source velocyto/bin/activate

pip install -U scvelo

pip install igraph # This will be used in the clustering later.
pip install louvain
```

# Executing scVelo: Using the existing umap

I plotted Oligo's UMAP in R, and now I want to plot the RNA velocity results on the same UMAP using scVelo software. I have already obtained the unspliced ​​expression matrix using velocityto. What should I do next?

This is a very classic requirement: **"R coordinates + Seurat clustering + Python (scVelo) rate analysis"**.

Since `scVelo` is based on Python, and your analysis is performed in R (Seurat), the core challenge lies in **how to seamlessly transfer the UMAP coordinates and cell annotations from R to Python**.

The current best practice workflow is: **R (export .h5ad) -> Python (load and merge .loom) -> scVelo analysis**.

Here are the detailed steps:

### Step 1: Prepare and export data in R

You need to convert the prepared Oligo Seurat object (containing UMAP coordinates and meta.data) to Python's `AnnData` format. The most commonly used package is `SeuratDisk`.

**R 代码：**

```{r}
library(Seurat)
library(SeuratDisk)

# env: dhlab-17 docker

sobj <- readRDS("002SC/004Seurat_out/002cell_type_anno/sobj.rds")
sobj.sub <- subset(sobj, anno.L3 %in% c("Oligo_0", "Oligo_1"))

# Assume your object is called oligo_seurat

# 1. Ensure your Default Assay is RNA and contains counts
DefaultAssay(sobj.sub) <- "RNA"

sobj.sub <- ScaleData(sobj.sub, features=rownames(sobj.sub))

# 2. This step is crucial: For scVelo to recognize UMAP, you need to ensure the reduction name is correct.

# Seurat is usually called 'umap', which will become 'X_umap' in Python after conversion, and this is fine.

# 3. Convert the Seurat object to an h5Seurat file

SaveH5Seurat(sobj.sub, filename = "002SC/004Seurat_out/002cell_type_anno/oligo_seurat.h5Seurat",
overwrite = T)

# 4. Convert h5Seurat to h5ad (Python AnnData format)
Convert("002SC/004Seurat_out/002cell_type_anno/oligo_seurat.h5Seurat", dest = "h5ad", overwrite=T)
```

---

### Step 2: Load and integrate the data in Python


```{python}
# env: python venv
# dir: optiplex, /storage/project/2022-10-14_mouse_spatial/scvelo

source velocyto/bin/activate

# scp lch@10.119.31.95:/storage/lch/project/2022-10-14_mouse_spatial/002SC/004Seurat_out/002cell_type_anno/oligo_seurat.h5ad ./

import scvelo as scv
import scanpy as sc
import anndata

adata1 = sc.read_h5ad("oligo_seurat.h5ad")

ldata_list = []
for sample in "AD_5622.1 AD_ctrl.1 WT_ctrl.1 AD_5622.2 AD_ctrl.2 WT_ctrl.2 AD_5622.4 AD_ctrl.4 WT_ctrl.4".split(): 
    ldata = sc.read("cellranger.out/" + sample + "/velocyto/" + sample + ".loom", cache=False)

    ldata.obs_names = [sample + "_" + x.split(':')[1].replace('x', '-1') for x in ldata.obs_names]
    
    cell_id_list = [i for i in adata1.obs_names if i.startswith(sample)]
    subset_adata = ldata[ldata.obs_names.isin(cell_id_list), :].copy()
    ldata_list.append(subset_adata)

for i, adata in enumerate(ldata_list):
    if not adata.var_names.is_unique:
        duplicates = adata.var_names[adata.var_names.duplicated()].unique()
        print(f"Index {i} contains duplicate gene names, for example: {duplicates[:5]}")

def remove_duplicate_vars(adata):
    return adata[:, ~adata.var_names.duplicated()].copy()

ldata_list_clean = [remove_duplicate_vars(a) for a in ldata_list]

merged_ldata = anndata.concat(ldata_list_clean, join="inner")

adata2 = scv.utils.merge(adata1, merged_ldata)

print(adata)

```

---

Step 3: Run RNA Velocity

```{python}
import numpy as np
import random
import os

seed = 42          
np.random.seed(seed)
random.seed(seed)
os.environ["PYTHONHASHSEED"] = str(seed)   

scv.pp.filter_and_normalize(adata2, min_shared_counts=20, n_top_genes=2000)
scv.pp.moments(adata2, n_pcs=30, n_neighbors=30)

scv.tl.recover_dynamics(adata2)
scv.tl.velocity(adata2, mode='dynamical')

scv.tl.velocity_graph(adata2)

scv.pl.velocity_embedding_stream(
    adata2, 
    basis='umap', 
    color=['anno.L3'], 
    title='Oligo RNA Velocity',  save="umap3.png", dpi=300
)
```

# scvelo: OPC+Oligo

## Step 1: Prepare and export data in R

```{r}

library(Seurat)
library(SeuratDisk)

sobj.sub <- readRDS("002SC/004Seurat_out/002cell_type_anno/Oligo/OPC_Oligo.rds")

DefaultAssay(sobj.sub) <- "RNA"

sobj.sub <- ScaleData(sobj.sub, features=rownames(sobj.sub))

SaveH5Seurat(sobj.sub, filename = "002SC/004Seurat_out/002cell_type_anno/Oligo/oligo_OPC.h5Seurat", 
             overwrite = T)

Convert("002SC/004Seurat_out/002cell_type_anno/Oligo/oligo_OPC.h5Seurat", dest = "h5ad", overwrite=T)
```

## Step 2: Loading and integrating data in Python

```{python}
# env:
source velocyto/bin/activate

# scp lch@10.119.31.95:/storage/lch/project/2022-10-14_mouse_spatial/002SC/004Seurat_out/002cell_type_anno/Oligo/oligo_OPC.h5ad ./

import scvelo as scv
import scanpy as sc
import anndata

adata = sc.read_h5ad("oligo_OPC.h5ad")

ldata_list = []
for sample in "AD_5622.1 AD_ctrl.1 WT_ctrl.1 AD_5622.2 AD_ctrl.2 WT_ctrl.2 AD_5622.4 AD_ctrl.4 WT_ctrl.4".split(): 
    ldata = sc.read("cellranger.out/" + sample + "/velocyto/" + sample + ".loom", cache=True)
    
    ldata.obs_names = [sample + "_" + x.split(':')[1].replace('x', '-1') for x in ldata.obs_names]
    
    cell_id_list = [i for i in adata.obs_names if i.startswith(sample)]
    subset_adata = ldata[ldata.obs_names.isin(cell_id_list), :].copy()
    ldata_list.append(subset_adata)

for i, adata in enumerate(ldata_list):
    if not adata.var_names.is_unique:
        duplicates = adata.var_names[adata.var_names.duplicated()].unique()
        print(f"Index {i} contains duplicate gene names, for example: {duplicates[:5]}")

def remove_duplicate_vars(adata):

    return adata[:, ~adata.var_names.duplicated()].copy()

ldata_list_clean = [remove_duplicate_vars(a) for a in ldata_list]

merged_ldata = anndata.concat(ldata_list_clean, join="inner")

adata2 = scv.utils.merge(adata, merged_ldata) # 

print(adata2)
```

### Step 3: Run RNA Velocity

```{python}
scv.pp.filter_and_normalize(adata2, min_shared_counts=20, n_top_genes=2000)
scv.pp.moments(adata2, n_pcs=30, n_neighbors=30)

scv.tl.recover_dynamics(adata2)
scv.tl.velocity(adata2, mode='dynamical')

scv.tl.velocity_graph(adata2)

scv.pl.velocity_embedding_stream(
    adata2, 
    basis='umap', 
    color=['anno.L3'], 
    title='Oligo OPC RNA Velocity',  save="umap4.png"
)
```
