
## data preprocessing

### sn data

```{r}
# env: dhlab-17
# dir: /home/mouse_ad/002SC
# sn data

library(SeuratDisk)
library(SingleCellExperiment)

sobj2 <- CreateSeuratObject(counts=sobj@assays$RNA@counts, 
                            meta.data=sobj@meta.data)

fp <- "./004Seurat_out/002cell_type_anno/sobj_for_deconvolution.h5Seurat"

SaveH5Seurat(sobj2, filename = fp)
# Creating h5Seurat file for version 3.1.5.9900
# Adding counts for RNA
# Adding data for RNA
# No variable features found for RNA
# No feature-level metadata found for RNA


Convert(fp, dest = "h5ad") 

# 运行成功：
# Validating h5Seurat file
# Adding data from RNA as X
# Adding counts from RNA as raw
# Transfering meta.data to obs

ls -ahl
# -rw-r--r--. 1 lch lch 1.9G Dec 20 22:21 sobj_for_deconvolution.h5ad

```

#### using pseudobulk

When performing deconvolution analysis on combined single-cell and spatial transcriptome data, is using a pseudobulk for the single-cell data better than using the original single-cell matrix?

This is a crucial and fundamental question in integrating single-cell and spatial transcriptome analysis. The answer is: for most spatial deconvolution tasks, using a pseudobulk is generally superior to using the original single-cell matrix directly, and is the recommended approach for current mainstream tools.

```{r}

setwd("/home/mouse_ad/002SC")

library(SeuratDisk)
library(Seurat)

# anno.L1
for (LEVEL in c("anno.L1", "anno.L2", "anno.L3")) {
  if (F) {
    LEVEL <- "anno.L3"
  }
  summed <- readRDS(paste0("004Seurat_out/004DEGs/DESeq/", LEVEL, "/summed.", LEVEL, ".rds"))
  meta2 <- data.frame(summed@colData[, c("orig.ident", LEVEL, "batch", "treatment")])
  row.names(meta2) <- paste0(meta2$orig.ident, "_", meta2[[LEVEL]])
  matrix2 <- assays(summed)$counts
  colnames(matrix2) <- row.names(meta2)
  sobj2 <- CreateSeuratObject(matrix2, meta.data=meta2)
  
  fp <- paste0("004Seurat_out/004DEGs/DESeq/", LEVEL, "/summed.", LEVEL, ".h5Seurat")
  SaveH5Seurat(sobj2, filename = fp)
  Convert(fp, dest = "h5ad") 
}
```


### st data

It seems that all you need to do is copy the h5ad file from the original data.

## run

```{bash}

# env: HPCC, conda (cell2loc_env)
# dir: /icto/user/yc27658/data-home/project/2022-10-14_mouse_spatial/cell2location # L1
# dir: /icto/user/yc27658/data-home/project/2022-10-14_mouse_spatial/cell2location.L2 # L2
# dir: /icto/user/yc27658/data-home/project/2022-10-14_mouse_spatial/cell2location.L3 # L3

srun -p gpu-short --gres=gpu:1 --mem=40G --time=01:00:00 --pty /bin/bash

conda activate cell2loc_env

python train_model.1.py

python view_step1.py

python train_model2.1.py

```

Using only one GPU, it takes approximately 45 minutes to run one sample. On HPCC, submit a single batch script, running only one sample at a time. This fully utilizes the GPU-short nodes, ensuring each sample completes its run while allowing the next script to run immediately after the previous one finishes.

Note that memory should be set to 40GB; setting it to 20GB will result in an insufficient memory warning.

Actually, you can schedule 3 tasks within 3 hours. HPCC limits a maximum of 5 tasks to running simultaneously, with 5 tasks queued.

```{bash}
#!/bin/bash
#SBATCH --job-name              cell2location
#SBATCH --partition             gpu-short
#SBATCH --nodes                 1
#SBATCH --tasks-per-node        1
#SBATCH --time                  3:00:00
#SBATCH --mem                   40G
#SBATCH --gres                  gpu:1
#SBATCH --output                ./log/cell2location.%j.out
#SBATCH --error                 ./log/cell2location.%j.err
#SBATCH --mail-type             ALL
#SBATCH --mail-user             yc27658@um.edu.com

source /etc/profile
source /etc/profile.d/modules.sh

#Adding modules
module add cuda/9.2.148
module add amber/18/gcc/4.8.5/cuda/9

ulimit -s unlimited

#Your program starts here
echo $CUDA_VISIBLE_DEVICES
python script/train_model2.1.py
```

```{bash}
sbatch script/submit-coral-gpu.sh.1
sbatch script/submit-coral-gpu.sh.2
sbatch script/submit-coral-gpu.sh.3
sbatch script/submit-coral-gpu.sh.4
sbatch script/submit-coral-gpu.sh.5
sbatch script/submit-coral-gpu.sh.6
```

## Results Summary

```{r}
# env: dhlab-17 docker

setwd("/home/mouse_ad/001ST")

library(Giotto)
gobj <- readRDS("001Giotto_out/testcombo.rds")

for (LEVEL in c("anno.L1", "anno.L2", "anno.L3")) {
  # LEVEL <- "anno.L1"
  
  cell.meta <- pDataDT(gobj)
  cell.meta.sub <- data.frame(cell.meta[, c("cell_ID", "cell_types", "list_ID")])
  colnames(cell.meta.sub) <- c("cell.id", "brain.region", "sample")
  
  # Combine data from several samples.
  samples <- unique(cell.meta$list_ID)
  names(samples) <- samples
  
  spot.anno.list <- list()
  for (sample in samples) {
    sample2 <- sub("_", ".", sample, fixed = T)
    in.file <- paste0("001Giotto_out/cell2location2/result_", 
                     LEVEL, "/cell2location_map_", 
                     sample2, "/", sample2, ".cell2loc.csv")
    data <- read.csv(in.file, row.names=1)
    cell.id <- paste0(sample, "-", row.names(data))
    data <- cbind(cell.id=cell.id, data[, 7:ncol(data)])
    spot.anno.list[[sample]] <- data
  }
  
  spot.anno.df <- data.frame(do.call(rbind, spot.anno.list))
  row.names(spot.anno.df) <- NULL
  spot.anno.df <- merge(spot.anno.df, cell.meta.sub, by="cell.id")
  
  end.col <- ncol(spot.anno.df) - 2
  total_abundance <- rowSums(spot.anno.df[, 2:end.col])
  spot.anno.prop <- cbind(spot.anno.df[, c("cell.id", "sample", "brain.region")], 
                              spot.anno.df[, 2:end.col]/total_abundance)
  
  write.csv(spot.anno.prop, paste0("001Giotto_out/cell2location2/", LEVEL, ".spot.anno.prop.csv"))
}

```